---
title: "Medicaid Expansion - Differences in Differences"
author: "Mason Turpin"
date: "March 6, 2019"
output: github_document
---

First, I read in the data. This data was given as part of an assignment in an Empirical Public Economics course taught by Dr. Sibley at UT-Austin. The assignment was to find out if there was a significant effect on individuals having insurance and individuals skipping healthcare due to cost after Medicaid was expanded in Ohio, and not expanded in Texas. Ohio was chosen since it was favorable to the parallel trends assumption essential to a DiD analysis.
```{r}
library(foreign)
library(tidyverse)

data_oh <- read.dta("data/DID_OH_ver12.dta")
```

Next, I consider two types of models to accomodate a Differences-in-Differences framework. The first is a linear probability model. It has the advantage of ease of interpretation, and often does well enough to make it a favorable choice. I also consider a logistic model, which will bind predictions between 1 and 0. This makes sense given our outcomes are binary variables. Our outcomes, again, are whether an individual has health insurance and whether they have skipped care due to cost. I append the fitted values for ease of plotting.
```{r}
model_plan <- lm(hlthpln1 ~ treat + post + treat_post + male + veteran3 + children +
              married + divorced + widowed + separated + never_married +
              log_income + edu_hs_grad + edu_college_attd + 
              edu_college_grad, data_oh)
summary(model_plan)

model_cost <- lm(medcost ~ treat + post + treat_post + male + veteran3 + children +
              married + divorced + widowed + separated + never_married +
              log_income + edu_hs_grad + edu_college_attd + 
              edu_college_grad, data_oh)
summary(model_cost)

data_oh$plan_fit <- model_plan$fitted.values
data_oh$cost_fit <- model_cost$fitted.values

#logistic
logi_plan <- glm(hlthpln1 ~ treat + post + treat_post + male + veteran3 + children +
                 married + divorced + widowed + separated + never_married +
                 log_income + edu_hs_grad + edu_college_attd + 
                 edu_college_grad, data = data_oh, family = "binomial")
summary(logi_plan)

logi_cost <- glm(medcost ~ treat + post + treat_post + male + veteran3 + children +
                                married + divorced + widowed + separated + never_married +
                                log_income + edu_hs_grad + edu_college_attd + 
                                edu_college_grad, data = data_oh, family = "binomial")
summary(logi_cost)

data_oh$logi_plan_fit <- logi_plan$fitted.values
data_oh$logi_cost_fit <- logi_cost$fitted.values
```

Last, I plot the results of our analysis. As you can see from each plot (not to mention the output above), there is a small but significant change in each outcome after the policy change.
```{r}
library(ggthemes)
library(scales)
data_oh$treat <- factor(data_oh$treat, levels = c(0, 1), labels = c("Texas", "Ohio"))
data_oh$year <- factor(data_oh$year)

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(hlthpln1_prop = sum(hlthpln1)/length(hlthpln1))) +
  geom_line(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion with Health Insurance", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Raw Values",
       caption = "The ACA went into full effect on January 1st, 2014. 
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(medcost_prop = sum(medcost)/length(medcost))) +
  geom_line(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion Skipping Care due to Cost", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Raw Values",
       caption = "The ACA went into full effect on January 1st, 2014. 
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(hlthpln1_prop = sum(plan_fit)/length(plan_fit))) +
  geom_line(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion with Health Insurance", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Fitted Values from Linear Probability Model",
       caption = "The ACA went into full effect on January 1st, 2014. 
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(medcost_prop = sum(cost_fit)/length(cost_fit))) +
  geom_line(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion Skipping Care due to Cost", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Fitted Values from Linear Probability Model",
       caption = "The ACA went into full effect on January 1st, 2014.
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(hlthpln1_prop = sum(logi_plan_fit)/length(logi_plan_fit))) +
  geom_line(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = hlthpln1_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion with Health Insurance", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Fitted Values from Logistic Model",
       caption = "The ACA went into full effect on January 1st, 2014. 
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()

ggplot(data = data_oh %>% 
         group_by(treat, year) %>% 
         summarise(medcost_prop = sum(logi_cost_fit)/length(logi_cost_fit))) +
  geom_line(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_point(aes(x = year, y = medcost_prop, group = treat, color = treat)) +
  geom_vline(xintercept = 3.5) +
  labs(x = "Year", y = "Population Proportion Skipping Care due to Cost", color = "State", 
       title = "Effect of Medicaid Expansion on Healthcare Trends", subtitle = "Fitted Values from Logistic Model",
       caption = "The ACA went into full effect on January 1st, 2014.
       Ohio chose to expand Medicaid while Texas did not.") +
  theme_classic() + scale_color_pander()
```